---
output: rmarkdown::github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

## smcsamplers

Implements SMC samplers to re-create the figures of the article entitled *An invitation to sequential Monte Carlo samplers*, written by Chenguang Dai, Jeremy Heng, Pierre E. Jacob, and Nick Whiteley, [available on arXiv at arxiv.org/abs/2007.11936](https://arxiv.org/abs/2007.11936).

This is not a general-purpose statistical software. This is just a collection of scripts intended to reproduce the figures of a paper. Use at your own risk!

The folder inst/README.R describes which scripts generate which figures.

### Installation

The package can be installed from R via:

```{r install, eval=FALSE}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("pierrejacob/smcsamplers")
```

It depends on the packages `Rcpp`, `RcppEigen`, `tidyverse`, `doParallel`, `doRNG`, `ggplot2`, `ggridges`, `ggthemes`, `lubridate`, `cobs`
which are all on CRAN.
